#Peer Assessment 1


##Loading and preprocessing the data

```{r assemble the data and do first plot, echo=TRUE}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp,method="curl")
activity <- read.csv(unz(temp, "activity.csv"),header=TRUE)
unlink(temp)
```

##Load useful packages

```{r load packages,echo=TRUE,results="hide"}
library(data.table)
library(dplyr)
library(knitr)
```

##Q1: What is the average number of steps taken per day?

```{r first problem, echo=TRUE}
totalsteps<-group_by(activity,date)
dailysum<-summarise(totalsteps,dailysteps=sum(steps),dailymean=mean(steps),dailymedian=median(steps))
meansteps<-as.integer(mean(dailysum$dailysteps,na.rm=TRUE))
mediansteps<-median(dailysum$dailysteps,na.rm=TRUE)
```

The mean of the total number of steps per day is `r meansteps`.

The median of the total number of steps per day is `r mediansteps`.

The histogram of the total number of steps taken each day:

```{r histogram,echo=FALSE,fig.height=4}
hist(dailysum$dailysteps,breaks=20,main="Frequency of Total Steps Per Day",xlab="total steps")
#hist(dailysum$dailysteps,freq=FALSE,breaks=50,xlim=c(0,22000),main="Density of Total Daily Steps",xlab="Total Daily Steps")
# dev.off()
```

##Q2: What is the average daily activity pattern?

```{r problem 2}
# plot2 find average daily activity pattern
# first make interval a factor and unfactor date
dailyint<-activity
dailyint$interval<-as.factor(dailyint$interval)
dailyint$date<-as.character(dailyint$date)
# group by interval
dailyint<-group_by(dailyint,interval)
intervalsum<-summarise(dailyint,intervalsteps=sum(steps,na.rm=TRUE),intervalmean=mean(steps,na.rm=TRUE))
# find the 5-minute interval that contains the maximum number of steps across all days
maxinterval<-subset(intervalsum,intervalsteps==max(intervalsteps),select=c(interval,intervalsteps))
maxnum<-maxinterval[1,1]
```

The time series showing the average steps by time interval:

```{r,echo=FALSE}
plot(intervalsum$interval,intervalsum$intervalmean,type="l",xlab="time intervals",ylab="average steps",main="average daily activity patterns")
```

###The interval with the maximum number of steps is `r maxnum`.

##Q3: What is the impact of missing values in the data?

###3.1: How many missing values are in the data?

```{r calculate missing values,echo=TRUE}
# identify number of rows with missing values
countna<-is.na(activity$steps)
countmissing<-as.matrix(table(countna))
countmissing<-countmissing[2]
```

The number of missing values is `r countmissing`.

###3.2 Missing values will be replaced with the mean steps by interval

```{r replacing missing values,echo=TRUE}
# replace NA values with steps mean by interval mean
# first join tables to assign interval mean to every reading
newactivity<-merge(dailyint,intervalsum,by.x="interval",by.y="interval",all=TRUE)
completeactivity<-newactivity
completeactivity$steps<-ifelse(is.na(completeactivity$steps),completeactivity$intervalmean,completeactivity$steps)
totalsteps2<-group_by(completeactivity,date)
dailysum2<-summarise(totalsteps2,dailysteps=sum(steps),dailymean=mean(steps),dailymedian=median(steps))
```

###3.3 The histogram of the total steps taken each day, with missing data filled in:

```{r plot 3,echo=FALSE}
hist(dailysum2$dailysteps,breaks=10,main="Total Steps Per Day Complete ",xlab="Total Steps Per Day",ylab="Frequency")
```

###3.4 How are the results (mean and median) different when missing values are imputed?

```{r new stats,echo=TRUE,results="hide"}
meansteps2<-as.integer(mean(dailysum2$dailysteps,na.rm=TRUE))
mediansteps2<-median(dailysum2$dailysteps,na.rm=TRUE)
```

Clearly there is virtually no difference when replacing missing values with imputed values.  

Old mean (missing values): `r meansteps`. New mean (imputed): `r meansteps2`.
Old median (missing values): `r mediansteps`. New median (imputed): `r mediansteps2`.

##Q4: Are there differences in activity patterns between weekdays and weekends?

```{r weekday diffs}
# create weekend and weekday factors
completeactivity$newdate<-as.Date(completeactivity$date)
completeactivity$day<-weekdays(completeactivity$newdate)
# group by interval
dailyint2<-completeactivity
dailyint2<-group_by(dailyint2,interval)
#create weekend/weekday factor variable
dailyint2$days<-as.factor(ifelse(dailyint2$day=="Saturday"|dailyint2$day=="Sunday","weekend","weekd"))
# summarize the steps by interval for each plot (weekend,weekday)
subset_we<-dailyint2$days=="weekend"
intervalsum_we<-summarise(dailyint2[subset_we,],mean_day_we=mean(steps))
subset_wd<-dailyint2$days=="weekd"
intervalsum_wd<-summarise(dailyint2[subset_wd,],mean_day_wd=mean(steps))
# merge files
intervalsum4<-merge(intervalsum_we,intervalsum_wd,by.x="interval",by.y="interval",all=TRUE)
```

The pattern of steps on weekends vs. weekdays shows significant differences:

```{r plot4,echo=FALSE,fig.height=8}
par(mfrow=c(2,1),mar=c(4,8,4,8),lty=3)
with(intervalsum4,{
  plot(interval,mean_day_we,main="weekend",type="n",ylab="number of steps")
  lines(interval,mean_day_we,type="l")
  plot(interval,mean_day_wd,main="weekday",type="n",xlab="interval")
  lines(interval,mean_day_wd,type="l")
})
```
